Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-10-07 Articles/2019-ShannonEntropy/Source/ShannonEntropyIsVariability.tex
Another pass at the language

A world without any variety is barely conceivable. It certainly would not admit scientific investigation. One of us recalls but has been unable to relocate a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. Whether he made the statement or not, it is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable
at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.

Without committing ourselves to Leibniz's overarching principle, our aim in this paper is to draw attention to the specific nature of the variety that can be observed in the universe and the unique suitability of Shannon's entropy (or information) formula $-\sum_ip_i\,\textrm{ln}\,p_i$ for the characterization of the variety. Indeed, as we will show in the next section, opportunities for application of Shannon's formula are ubiquitous both on the earth and in the universe at large. PRECISELY BECAUSE OF THIS GENERALITY WE WILL OFTEN USE THE EXPRESSION SHANNON `FORMULA' RATHER THAN `ENTROPY' OR `INFORMATION'.

@@

This leads us to propose a different characterization of Shannon's formula that is both more general and more appropriate. Indeed, we will argue that the formula is a measure of the \emph{variability within a distribution of things of a given kind}. Moreover, the formula has a very wide range of applicability because the universe contains many different kinds of things that, through having common attributes, can be recognized as belonging together despite the fact that their attributes are expressed to different degrees or in different ways. To give a first example, the girls in a school can differ in the length and colour of their hair. More examples will follow in the next section. In this paper we do not ask why the universe sustains the variety that it does. We take it as a given.

NEW PARAGRAPH

@@

\vspace{.5in}

TODO: Introduction and Conclusions still need to be done

Current disputed status of Shannon entropy as a physical object.

Immense regularity in nature leads to categorization of elements intro distributions. Stress that these distributions may be of different nature? I.e.: statistical distributions of elements that exist at the same time, probability distributions of events that may happen given similar circumstances, credence distributions of expectation for an agent that something will happen. May be better to have a another word instead of "probability" when talking about the elements of the distribution. Maybe "weight" or "fraction"?

A natural question is to ask how much variability an element has within a distribution. Tempted to just count the number of cases over which the distribution is defined. But the shape also affects the variability (e.g. a distribution concentrated in few values means less variability of the elements than one equally distributed).

In this paper we will show that $- \sum p_i \log p_i$ is the most natural form to describe the variability of the elements of a distribution.

(NOTE: defining the concept like this is actually broader than Shannon entropy, which assumes source/destination/message.)

@@

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated with the bins are non-numerical (e.g. types of galaxies, words), statistical quantities are not well defined. Second, relabeling the values (e.g. switching names, switching units, non-linearly changing coordinates) does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require $H=H(p_i)$.\footnote{Note that for continuous quantities the weights are densities and are affected by the choice of $Q$: the unit is required to specify the numeric value (e.g. $1 \% / m$) and this will change under unit transformation (e.g. $1000 \% / Km$). Therefore, as the weights $p_i$ themselves depend on the unit, the indicator $H(p_i)$ will in general depend on the choice of $Q$. Though this is source of additional confusion, for the purpose of defining a measure of variability it does not change things conceptually. We will discuss the problem in connection to entropies in thermodynamics and statistical mechanics, since coordinate invariance is paramount in physics.}

@@

As we noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. What we would like to say is that the variability of the overall distribution $H_T = H_C + p_M H_M$ will be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$.

@@

The Shannon variability, then, is a clear and important attribute of a distribution: it quantifies the variability of the elements within the distribution. It quantifies the variety of values one finds within the distribution. It is true that the choice of the elements and binning are not fixed, it is true that the meaning of the weights depends on what the distribution is describing, and therefore that the precise meaning of this variability is context dependent. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope a probability and so on. And like a real number represents a quantity, transcending what is physically quantified, the Shannon variability represents the variability of the elements\footnote{We again stress that out elements are physical and therefore posses some property beyond the bare `definite and separate' of Cantor's definition, which was perfectly suited to study numbers. Without these extra attributes, which are `washed out' by said definition, we would not be able to create a distribution in the first place.} in a distribution, transcending what the distribution physically signifies.

@@

Unfortunately, entropy [CAN'T REALLY CHANGE IT HERE] in general is often associated with vague characterizations like the the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending by the authors[CITATION NEEDED], which leads to confusion and misunderstanding.  The characterization of the Shannon variability measure we have given, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability we find in his vocabulary? All these questions are interesting, all these questions are profound and the answer to all these questions are expressed through just one thing: the Shannon variability measure.

@@

We can gain insight with the following example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands grouped by their respective species. Suppose we pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

@@

The Shannon variability of the distribution, then, has nothing to do with the information the distribution itself holds. The distribution is not what is being encoded. The variability is the information needed to go from the distribution, which is given, to an individual element. It's really the information gap from the population to an element. That is why some people say the entropy [CAN'T CHANGE IT HERE] is ``lack of information", because it's the information the distribution does not have. But, again, this is just missing the point of what is being described. If one is not interested in identifying elements there is no ``missing information".

@@

The first question is straightforward to answer. The macroscopic variables of a system (e.g. volume, pressure, energy) do not give a full description of its parts. If we fix a macrostate and then measure the state of all its microscopic constituents, we will not get the same answer every time. This true in the case of non equilibrium (e.g. we suddenly compress a gas pushing a piston) and of equilibrium (e.g. we give enough time to equilibrate). The choice of a macrostate, then, identifies a set of possible microstates that are compatible with it. It is then valid to ask the question: what is the variability of microstates one can find within a given macrostate? This is answered by the Shannon variability, which gives us the number of bits required to identify a microstate given the macrostate. Since the macrostates fully defines the distribution of possible microstates, this distribution does not depend on the particular way the macrostate came to pass. It is path independent: it is a state variable. In short, as a macrostate is a distribution over microstates, it will admit as a state variable the Shannon variability, regardless of whether we know the actual distribution.

Note, though, that the thermodyanmic entropy is defined differently. It is defined over a succession of equilibrium states by noting that the integral of the heat exchanged divided by the temperature is path independent. One then chooses a particular state $A$, assigns it an arbitrary value $S_A$ and the entropy for all other states is given by the integral $S(B) = S_A + \int_A^B \frac{dQ}{T}$. At this point, it is not apparent why these two quantities defined independently would be related. What we can say, though, is that all quantities that increase over irreversible processes, are maximized at equilibrium and that are additive for indipendent systems are precisely the quantities that are linear functions of entropy.[CITE AXIOMATIC WORK] A linear function of entropy, though, simply means a redefinition of $S_A$ and a change of units for either $dQ$ or $T$. Therefore if we can show that the Shannon variability possesses those three properties than it is equivalent to the entropy up to a change in units, which brings us to the second question.

@@

The Shannon variability, then, has all the right properties to establish a conceptually meaningful connection with the thermodynamic/statistical mechanics entropy. It is indeed an independent concept but when applied to macrostates yields a state variable with the right conceptual characteristics. There are still some mathematical details that need to be worked out to have a complete picture. These details don't change the overall intuitive picture, but they present yet again opportunities for confusion.

The first is the role of the Boltzmann constant $k_B$, which is a conversion factor between the Shannon variability. $S = - \sum p_i \log(p_i)$ and the Gibbs entropy $S_G = - k_B p_i \sum \log(p_i)$. This conversion constant should not surprise us, since we only established a linear relationship between variability and entropy. The question is whether this conversion itself is has a physical meaning, or it is also simply due to a choice of units. The standard relationship between temperature, internal energy and entropy is $T = \frac{\partial U}{\partial S}$. Historically, temperature and internal energy are the quantities that were defined first. If we measure temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. If we changed units, $k_B$ would change accordingly.

@@

Alternatively, a microstate can be represented as a point over $N$ copies of phase space, sometimes referred to as $\Gamma$-space. It is on this space that the Gibbs entropy is defined and where thermodynamic entropy happens to coincide with the Shannon entropy. Since this is a probability distribution, the microstate corresponds to a single element within the distribution defined by the macrostate. Yet, one has to be careful of two things. The first is that permutations of particles within $\Gamma$-space lead to different points which should not lead to different microstates. This leads to the widely known problem of overcounting which needs to be addressed in the standard way. The second problem is that we should not think of the particles as literally pointlike: they would correspond to delta Dirac distributions over phase-space which have minus infinite Shannon entropy. It is more appropriate, both mathematically and conceptually, to think of indistinguishable particles as identical distributions over $\mu$-space, each contributing the same amount of Shannon entropy.\footnote{Setting to $-\log h$ the entropy corresponding to each degree of freedom of these single particle distributions is a natural way to incorporate the effects of the uncertainty principle.}

@@

What happens in phase space is that, under coordinate transformation $\hat{q}^i = \hat{q}^i(q^j)$, the $d\hat{q}^i = \frac{\partial \hat{q}^i}{\partial q^j } dq^j$ vary like vector components while $d\hat{p}_i = \frac{\partial q^j}{\partial \hat{q}^i } dp_j$ vary like covector components. Therefore the areas $d\hat{q}^i d\hat{p}_i = dq^j dp_j$ remain the same. The Jacobian determiants are unitary and the Shannon entropy is invariant. It is not the invariance of the areas over Hamiltonian evolution that is important: it is the invariance over coordinate transformations. Invariance over time would be insignificant if we first didn't have a quantity that was observer independent.

This means that we always have to consider distributions over the full phase-space with the appropriate conjugate coordinates if we want to use $-\rho \log \rho$. If we use distributions over position and velocities or over momentum only, the expression we changed for those particular variables in those coordinate systems.

These technical details are important to not get confused, but they do not alter the conceptual point. The Shannon entropy and the thermodynamic entropy, though describing different things, are conceptually related. The variability of the state of the parts within the state of the whole is what the Shannon entropy quantifies if applied to the macrostates of thermodynamics and the microstates of statistical mechanics. This variability increases during irreversible evolution, more microstates converge to the same macrostate until equilibrium is reached. The variability of one system does not affect the variability of an independent one and therefore it simply sums. This is not a coincidence: this is what thermodynamic entropy ultimately is about.

>>>>

A world without any variety is barely conceivable. It certainly would not admit scientific investigation. We recall, though we have been unable to relocate, a sentence in one of Leibniz's many writings to the effect that reality \emph{is} variety. Whether he made the statement or not, it is consistent with his view that any contingent thing is defined by a collection of attributes, which necessarily must be distinguishable
at some epistemic level. Paragraph 58 of Leibniz's \emph{Mondadology} asserts that the principle underlying the universe is that it should realize ``as much variety as possible, but with the greatest order possible''.

Without committing ourselves to Leibniz's overarching principle, our aim in this paper is to draw attention to the specific nature of the variety that can be observed in the universe and the unique suitability of Shannon's entropy (or information) formula $-\sum_ip_i\,\textrm{ln}\,p_i$ for the characterization of the variety. Indeed, as we will show in the next section, opportunities for application of Shannon's formula are ubiquitous both on the earth and in the universe at large.

@@

This leads us to propose a different characterization of Shannon's formula that is both more general and more appropriate. Indeed, we will argue that the formula is a measure of the \emph{variability of the values within a distribution of a given set of elements} and we will use the expression Shannon ``variability'', rather than ``entropy'' or ``information''. Moreover, the formula has a very wide range of applicability because the universe contains many different kinds of things that, through having common attributes, can be recognized as belonging together despite the fact that their attributes are expressed to different degrees or in different ways. To give a first example, the girls in a school can differ in the length and colour of their hair. More examples will follow in the next section. In this paper we do not ask why the universe sustains the variety that it does. We take it as a given.

@@

@@

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated with the bins are non-numerical (e.g. types of galaxies, words), statistical quantities are not well defined. Second, relabeling the values (e.g. switching names, switching units, non-linearly changing coordinates) does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require $H=H(p_i)$.\footnote{Note that for continuous quantities the weights are densities and are affected by the choice of $Q$: the unit is required to specify the numeric value (e.g. $1 \% / m$) and this will change under unit transformation (e.g. $1000 \% / Km$). Therefore, as the weights $p_i$ themselves depend on the unit, the indicator $H(p_i)$ will in general depend on the choice of $Q$. Though this is source of additional confusion, for the purpose of defining a measure of variability it does not change things conceptually. We will discuss the problem in connection to entropies in thermodynamics and statistical mechanics, since coordinate invariance is of paramount importance in physics.}

@@

As we noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. What we would like to say is that the variability of the overall distribution $H_T = H_C + p_M H_M$ will be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity linear when combining two independent distributions.

@@

The Shannon entropy, which we will call Shannon variability, is a clear and important attribute of a distribution: it quantifies the variability of the elements within the distribution. It quantifies the variety of values one finds within the distribution. It is true that the choice of the elements and binning are not fixed, it is true that the meaning of the weights depends on what the distribution is describing, and therefore that the precise meaning of this variability is context dependent. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope a probability and so on. And like a real number represents a quantity, transcending what is physically quantified, the Shannon variability represents the variability of the elements\footnote{We again stress that out elements are physical and therefore posses some property beyond the bare `definite and separate' of Cantor's definition, which was perfectly suited to study numbers. Without these extra attributes, which are `washed out' by said definition, we would not be able to create a distribution in the first place.} in a distribution, transcending what the distribution physically signifies.

@@

Unfortunately, entropy in general is often associated with vague characterizations like the the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending by the authors[CITATION NEEDED], which leads to confusion and misunderstanding.  The characterization of the Shannon variability measure we have given, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability we find in his vocabulary? All these questions are interesting, all these questions are profound and the answer to all these questions are expressed through just one thing: the Shannon variability.

@@

We can gain insight with the following example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands binned by their respective species. Suppose we pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

@@

The Shannon variability of the distribution, then, has nothing to do with the information the distribution itself holds. The distribution is not what is being encoded. The variability is the information needed to go from the distribution, which is given, to an individual element. It's really the information gap from the population to an element. That is why some people say the entropy is ``lack of information", because it's the information the distribution does not have. But, again, this is just missing the point of what is being described. If one is not interested in identifying elements there is no ``missing information".

@@

The first question is straightforward to answer. The macroscopic variables of a system (e.g. volume, pressure, energy) do not give a full description of its parts. If we fix a macrostate and then measure the state of all its microscopic constituents, we will not get the same answer every time. This true in the case of non equilibrium (e.g. we suddenly compress a gas pushing a piston) and of equilibrium (e.g. we give enough time to equilibrate). The choice of a macrostate, then, identifies a set of possible microstates that are compatible with it. It is then valid to ask the question: what is the variability of microstates one can find within a given macrostate? This is answered by the Shannon variability, which gives us the number of bits required to identify a microstate given the macrostate. Since the macrostates fully define the distribution of possible microstates, this distribution does not depend on the particular way the macrostate came to pass. It is path independent: it is a state variable. In short, as a macrostate is a distribution over microstates, it will admit as a state variable the Shannon variability, regardless of whether we know the actual distribution.

Note, though, that the thermodyanmic entropy is defined differently. It is defined over a succession of equilibrium states by noting that the integral of the heat exchanged divided by the temperature is path independent. One then chooses a particular state $A$, assigns it an arbitrary value $S_A$ and the entropy for all other states is given by the integral $S(B) = S_A + \int_A^B \frac{dQ}{T}$. At this point, it is not apparent why these two quantities defined independently would be related. What we can say, though, is that all quantities that increase over irreversible processes, are maximized at equilibrium and that are additive for indipendent systems are precisely the quantities that are linear functions of entropy.[CITE AXIOMATIC WORK] A linear function of entropy, though, simply means a redefinition of $S_A$ and a change of units for either $dQ$ or $T$. Therefore if we can show that the Shannon variability possesses those three properties than it is equivalent to the entropy up to a change in units, which would answer our second question.

@@

The Shannon variability, then, has all the right properties to establish a conceptually meaningful connection with the thermodynamic/statistical mechanics entropy. It is a state variable that represents the variability of microstates within a macrostate. It increases during irreversible processes, reaches the maximum at equilibrium and it is additive for independent systems. It is indeed an independent concept but when applied to macrostates yields a state variable with the right conceptual characteristics. There are still some mathematical details that need to be worked out to have a complete picture. These details don't change the overall intuitive picture, but they present yet again opportunities for confusion.

The first is the role of the Boltzmann constant $k_B$, which is a conversion factor between the Shannon variability. $H = - \sum p_i \log(p_i)$ and the Gibbs entropy $S_G = - k_B p_i \sum \log(p_i)$. This conversion constant should not surprise us, since we only established a linear relationship between variability and entropy. The question is whether this conversion constant itself has a physical meaning, or it is also simply due to a choice of units. The standard relationship between temperature, internal energy and entropy is $T = \frac{\partial U}{\partial S}$. Historically, temperature and internal energy are the quantities that were defined first. If we measure temperature in Kelvin and energy in Joules, the above relationship forces us to measure entropy in Joules/Kelvin. If we changed units, $k_B$ would change accordingly.

@@

Alternatively, a microstate can be represented as a point over $N$ copies of phase space, sometimes referred to as $\Gamma$-space. It is on this space that the Gibbs entropy is defined and where thermodynamic entropy happens to coincide with the Shannon entropy. Since this is a probability distribution, the microstate corresponds to a single element within the distribution defined by the macrostate. Yet, one has to be careful of two things. The first is that permutations of particles within $\Gamma$-space lead to different points which should not lead to different microstates. This leads to the widely known problem of overcounting which needs to be addressed in the standard way. The second problem is that we should not think of the particles as literally pointlike: they would correspond to delta Dirac distributions over phase-space which have minus infinite Shannon entropy. It is more appropriate, both mathematically and conceptually, to think of indistinguishable particles as identical peaked distributions over $\mu$-space, each contributing the same amount of Shannon entropy.\footnote{Setting to $-\log h$ the entropy corresponding to each degree of freedom of these peaked distributions is a natural way to incorporate the effects of the uncertainty principle.}

@@

What happens in phase space is that, under coordinate transformation $\hat{q}^i = \hat{q}^i(q^j)$, the $d\hat{q}^i = \frac{\partial \hat{q}^i}{\partial q^j } dq^j$ vary like vector components while $d\hat{p}_i = \frac{\partial q^j}{\partial \hat{q}^i } dp_j$ vary like covector components. Therefore the areas $d\hat{q}^i d\hat{p}_i = dq^j dp_j$ remain the same. The Jacobian determiants are unitary and the Shannon entropy is invariant. This brings us to another point of confusion: it is often argued that phase space areas are important because they are conserved under Hamiltonian evolution. While they are indeed conserved, that is not the reason why they are important. It is their invariance under coordinate transformation that makes them physically significant. In fact, conservation over time would not even be well defined if the quantity was not invariant and therefore have different values depending on the observer.

This means that we always have to consider distributions over the full phase-space with the appropriate conjugate coordinates if we want to use $-\rho \log \rho$. If we use distributions over position and velocities or over momentum only, the expression would change for those particular variables in those coordinate systems. This can be, again, a source of confusion.

These technical details are important to use the Shannon variability correctly, but they do not alter the conceptual point. The Shannon variability, the thermodynamic entropy and the Gibbs entropy, though describing different things, are conceptually related. The variability of the state of the parts within the state of the whole is what the Shannon variability quantifies if applied to the macrostates of thermodynamics and the microstates of statistical mechanics. This variability increases during irreversible evolution, more microstates converge to the same macrostate until equilibrium is reached, where the variability is maximized. The variability of one system does not affect the variability of an independent one and therefore it simply sums. This is not a coincidence: this is what thermodynamic entropy ultimately is about.
